Higher-Order unification for free!

### 2024-05-02 paper/main.tex
proofreading 1, 2 and 3

\subsection{The problem: logic-program simulation}
We represent a logic program \emph{run} in \Fo{} as
a sequence of $n$ \emph{steps}. Each step consists in
unifying two terms, $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$, taken from the list of all unification
problems \foUnifPb.
% \footnote{If the same rule is used multiple times in a run we
% consider as many copies as needed of the terms composing the
% rules, with fresh unification variables each time. Also, we do not
% model backtracking directly, but rather the fact that a step in a run may
% fail, and in that case the entire run is said to fail. A regular logic program
% that succeeds is modeled as a many runs, with possibly overlapping starts,
% where at least one run does not end in a failure. }
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_n$, which is the result of the
logic program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{ostep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{mstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{orun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{mrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFx}[3]{\ensuremath{\frun(#1,#2) \mapsto #3}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHx}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{n}{\rho}
\deff
\runFD{\foUnifPb}{n}
\end{array}
$$

\noindent
In order to simulate a logic program we start by compiling
\foUnifPb to \hoUnifPb, i.e.
each \Fo{}-term $s \in \foUnifPb$ is translated to a \Ho{}-term $t \in \hoUnifPb$.
We write this translation as $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produces a variable map $m$ and a list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.
We write $\foUnifPb_p = \{~ \foUnifPb_{p_l},\, \foUnifPb_{p_r} ~\}$
and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.

@@

\begin{proposition}[Fidelity recovery]\label{prop:fidelity-recovery} 
In the context of \hrun, if 
$\rho_{p-1} \foUnifPb_{p} \in \llambda$ 
(even if $\;\foUnifPb_{p} \not\in \llambda$)
then
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p} \Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p}
$$
\end{proposition}

@@

\paragraph{Backtracking} We omit it from our model of a logic-program execution
since it plays a very minor role, orthogonal to higher-order unification.
We point out that each \emph{run} corresponds to a (proof search) branch in the
logic program that either fails at some point, or succeeds. A computation that
succeeds by backtracking, exploring multiple branches, could be
modeled as a set of runs with a (possibly non-empty) common prefix.

\subsection{The solution (in a nutshell)}
\label{sec:nutshell}
A term $s$ is compiled to a term $t$ where every
``problematic'' subterm $e$ is replaced by a fresh unification variable $h$
with an accessory \emph{link} that represents a suspended unification problem
$h \Ue e$. As a result, \Ue is ``well behaved'' on $t$, in the sense that
it does not contradict \Eo as it would otherwise do on the
``problematic'' sub-terms.\\
We now define ``problematic'' and ``well behaved'' more formally
We use the $\Diamond$ symbol since it stands for ``possibly'' in modal logic
and all problematic terms are characterized by some ``uncertainty''.

\begin{definition}[\maybebeta]\label{def:maybebeta}
  \maybebeta is the set of terms of the form $X \appsep x_1 \ldots x_n$
  such that $x_1 \ldots x_n$ are distinct names (of bound variables).
\end{definition}

@@

\begin{definition}[\notllambda]\label{def:notllambda}
  \notllambda is the set of terms of the form $X \appsep t_1 \ldots t_n$
  such that $t_1 \ldots t_n$ are not distinct names.
\end{definition}

@@

\noindent
In other words the compiler outputs terms in \wellb, even if its
input is not. And if two terms can unify in the source language
then their images also unify in the target language. 
Note that the property holds for any substitution, $\rho$ could be given by an
oracle and/or not necessarily be a most general one: when applied to terms in \wellb,
\Ue{} does not contradict \Eo. Note that a compiler that
translate all terms to unification variables would satisfy this
property, but wound certainly fail to verify \fullref{prop:fidelity}.

@@

\noindent
Proposition \ref{prop:w-preservation} is key to proving \cref{prop:simulation,prop:fidelity}.
Informally, it says that the problematic terms moved on the side by the compiler
are not reintroduced by \hstep, hence \Ue{} can continue to operate properly.
In \cref{sec:simulation,sec:eta,sec:beta}
we describe how to %the compiler
recognize terms in \maybebeta, \maybeeta and
\notllambda and how \progress takes care of them preserving \wellb
and ensuring \cref{prop:simulation,prop:fidelity,prop:fidelity-recovery}.

@@

\subsubsection{Scope check}\label{sec:sc}
The predicate \elpiIn{scope-check} walks through all links. 
Each link of the form \linketaM{\Gamma}{X_{x_1\ldots x_n}}{t} is replaced by
\linketaM{{x_1\ldots x_n}}{X_{x_1\ldots x_n}}{t'} whenever $\{x_1\ldots x_n\} \subset \Gamma$.
The term $t' = \lambda x.Y_{{x_1\ldots x_n},x}$ is obtained after
executing $\lambda x.Y_{{x_1\ldots x_n}~x} \Ue t$. This unification
ensures that $t'$ cannot contain variables in $\Gamma / \{x_1\ldots x_n\}$,
and therefore grants fidelity.
In fact, a link represents a suspended unification problem between a unification variable $X$
and a ``problematic'' term $t$. This unification is possible iff 
the free variables in $t$ are also in the scope of $X$.

\subsubsection{Occur check}\label{sec:oc}
Since compilation moves problematic terms out of the sight of \Ue{},
that procedure can only perform a partial occur check. For example, the
unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
but should be disallowed if $\linkStore$ contains a link like
$\linketaM{}{Y}{\lambda z.X_z}$: we don't know yet if $Y$ will feature
a lambda in head position, but we surely know it contains $X$, hence,
% $f~Y$ and 
that fails the occur check.
The procedure \elpiIn{occur-check-links} is in charge of
performing this check that is needed in order to
guarantee \fullref{prop:fidelity}.

@@

\noindent
So far the compiler is very basic. It does not really
enforce that the terms passed to \hstep are in \wellb,
and indeed makes no use of the higher-order capabilities
of the meta language (all generated
variables have an empty scope).
Still, we can prove that \Uo is a good
``first-order'' unification algorithm if the input
already happens to be in \wellb.
Later, when the compiler will unsure \fullref{prop:w-enforcing}
and
the proof will be refined to cover for the new cases.

\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow}
  The implementation of \Uo above is a good unification
  for \Eo
  (as per \cref{prop:good-unif})
  in the domain \wellb.
\end{lemma}
\begin{proof}[Proof sketch]
In this setting, \Ee is as strong as \Eo on ground terms and reflexivity for
unification variables. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
<<\elpiIn{fuva X ~\Uo~s}>>. In this case, after compilation, we have
<<\elpiIn{uva Y [] ~\Ue~t}>>
that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{proof}[Proof sketch]
  Let $\foUnifPb_i$ be the first problem in \foUnifPb, such that it is not in \llambda and le
  $\sigma$ be the substitution obtained solving
  $\hoUnifPb_1\ldots\hoUnifPb_{i-1}$. After the execution of $\hstep_{i-1}$,
  each variable $X$ corresponding to a \maybeeta subterm in the original \Fo 
  unification problem, is instantiated if it is known for sure
  that $X$ is no more a \maybeeta. If $\sigma\hoUnifPb_i$ is in \llambda, then 
  % each variable that the
  % compiler used to replace a ``problematic'' term is assigned and hence
  by \cref{def:progBetaLL,def:progBetaRH} the associated \linkbeta
  is solved and removed. Note that, in this case, all
  calls to \Ue{} are between terms in \llambda and by \cref{th:fidelity-beta}
  fidelity is guaranteed. If $\sigma\hoUnifPb$ is still in \notllambda,
  then by \cref{def:progBetaFail} unification fails as the
  corresponding unification in \Fo{} would.
\end{proof}

>>>>

\subsection{The problem: logic-program simulation}
We represent a logic program \emph{run} in \Fo{} as
a sequence of $n$ \emph{steps}. Each step $p$ consists in
unifying two terms, $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$, taken from the list of all unification
problems \foUnifPb.
We write $\foUnifPb_p = \{~ \foUnifPb_{p_l},\, \foUnifPb_{p_r} ~\}$
and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.
% \footnote{If the same rule is used multiple times in a run we
% consider as many copies as needed of the terms composing the
% rules, with fresh unification variables each time. Also, we do not
% model backtracking directly, but rather the fact that a step in a run may
% fail, and in that case the entire run is said to fail. A regular logic program
% that succeeds is modeled as a many runs, with possibly overlapping starts,
% where at least one run does not end in a failure. }
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_n$, which is the result of the
logic program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{{\mathrm{step}}_o}\xspace}
\newcommand{\hstep}{\ensuremath{{\mathrm{step}}_m}\xspace}
\newcommand{\frun}{\ensuremath{{\mathrm{run}}_o}\xspace}
\newcommand{\hrun}{\ensuremath{{\mathrm{run}}_m}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFx}[3]{\ensuremath{\frun(#1,#2) \mapsto #3}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHx}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{n}{\rho}
\deff
\runFD{\foUnifPb}{n}
\end{array}
$$

\noindent
In order to simulate a logic program we start by compiling
\foUnifPb to \hoUnifPb, i.e.
each \Fo{}-term $s \in \foUnifPb$ is translated to a \Ho{}-term $t \in \hoUnifPb$.
We write this translation as $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produces a variable map $m$ and a list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the substitution, that is written
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.

@@

\begin{proposition}[Fidelity recovery]\label{prop:fidelity-recovery} 
In the context of \hrun, if 
$\rho_{p-1} \foUnifPb_{p} \subseteq \llambda$ 
(even if $\;\foUnifPb_{p} \not\subseteq \llambda$)
then
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p} \Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p}
$$
\end{proposition}

@@

\paragraph{Backtracking} We omit it from our model of logic-program execution
since it plays a very minor role, orthogonal to higher-order unification.
We point out that each \emph{run} corresponds to a (proof search) branch in the
logic program that either fails at some point, or succeeds. A computation that
succeeds by backtracking, exploring multiple branches, could be
modeled as a set of runs with a (possibly non-empty) common prefix.

\subsection{The solution (in a nutshell)}
\label{sec:nutshell}
A term $s$ is compiled to a term $t$ where every
``problematic'' subterm $e$ is replaced by a fresh unification variable $h$
with an accessory \emph{link} that represents a suspended unification problem
$h \Ue e$. As a result, \Ue is ``well behaved'' on $t$, in the sense that
it does not contradict \Eo as it would otherwise do on the
``problematic'' sub-terms.\\
We now define ``problematic'' and ``well behaved'' more formally.
We use the $\Diamond$ symbol since it stands for ``possibly'' in modal logic
and all problematic terms are characterized by some uncertainty.

\begin{definition}[\maybebeta]\label{def:maybebeta}
  \maybebeta is the set of terms of the form $F \appsep x_1 \ldots x_n$
  such that $x_1 \ldots x_n$ are distinct names (of bound variables).
\end{definition}

@@

\begin{definition}[\notllambda]\label{def:notllambda}
  \notllambda is the set of terms of the form $F \appsep t_1 \ldots t_n$
  such that $t_1 \ldots t_n$ are not distinct names.
\end{definition}

@@

\noindent
In other words the compiler outputs terms in \wellb even if its
input is not. And if two terms can be unified in the source language
then their images can also be unified in the target language. 
Note that the property holds for any substitution, not necessarily a most
general one: when applied to terms in \wellb
the meta-language unification 
\Ue{} agrees with the object-language equality \Eo. Note that a compiler that
translate all terms to unification variables would satisfy this
property, but wound certainly fail to verify \cref{prop:fidelity}.

@@

\noindent
Proposition \ref{prop:w-preservation} is key to proving \cref{prop:simulation,prop:fidelity}.
Informally, it says that the problematic terms moved on the side by the compiler
are not reintroduced by \hstep, hence \Ue{} can continue to operate properly.
In \cref{sec:llam,sec:eta,sec:beta}
we describe how to %the compiler
recognize terms in \maybebeta, \maybeeta and
\notllambda and how \progress takes care of links and how it preserves \wellb
and ensures \cref{prop:simulation,prop:fidelity,prop:fidelity-recovery}.

@@

\subsubsection{Scope check}\label{sec:sc}
The predicate \elpiIn{scope-check} 
replaces each
link of the form \linketaM{\Gamma}{X_{x_1\ldots x_n}}{t} with
a link
\linketaM{{x_1\ldots x_n}}{X_{x_1\ldots x_n}}{t'} whenever $\{x_1\ldots x_n\} \subset \Gamma$.
The term $t' = \lambda x.Y_{{x_1\ldots x_n},x}$ is obtained after
executing $\lambda x.Y_{{x_1\ldots x_n}~x} \Ue t$ using a fresh $Y$.
This unification
ensures that $t'$ cannot contain variables in $\Gamma / \{x_1\ldots x_n\}$,
and if it fails then \progress hence \hstep fails.
In turn this grants fidelity in the case where the \lhs of an \linketa
is pruned of a name that occurs (rigidly) in $t$:
links represent suspended unification problems
\emph{that may succeed}.

\subsubsection{Occur check}\label{sec:oc}
Since compilation moves problematic terms out of the sight of \Ue{},
that procedure can only perform a partial occur check. For example, the
unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
but should be disallowed if $\linkStore$ contains a link like
$\linketaM{}{Y}{\lambda z.X_z}$: we don't know yet if $Y$ will feature
a lambda in head position, but we surely know it contains $X$.
The procedure \elpiIn{occur-check-links} is in charge of
performing this check that is needed in order to
guarantee \fullref{prop:fidelity}.

@@

\noindent
So far the compiler is very basic. It does not really
enforce that the terms passed to \hstep are in \wellb,
and indeed makes no use of the higher-order capabilities
of the meta language (all generated
variables have an empty scope).
Still, we can prove that \Uo is a good
``first-order'' unification algorithm if the input
already happens to be in \wellb.
Later, when the compiler will ensure \fullref{prop:w-enforcing}
and
the proof will be refined to cover for the new cases.

\begin{theorem}[Properties of \Uo in \wellb]\label{prop:uniffow}
  The implementation of \Uo above is a good unification
  for \Eo
  (as per \cref{prop:good-unif})
  in the domain \wellb.
\end{theorem}
\begin{proof}[Proof sketch]
In this setting, \Ee is as strong as \Eo on ground terms.
What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$,
we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, the only interesting case is
<<\elpiIn{fuva X ~\Uo~s}>>. In this case, after compilation, we have
<<\elpiIn{uva Y [] ~\Ue~t}>>
that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{proof}[Proof sketch]
  Let $\foUnifPb_i$ be the first problem in \foUnifPb, such that it is not in \llambda and le
  $\sigma$ be the substitution obtained solving
  $\hoUnifPb_1\ldots\hoUnifPb_{i-1}$. After the execution of \hstep
  for the $i-1$ time,
  each variable $X$ corresponding to a \maybeeta subterm in the original \Fo 
  unification problem, is instantiated if it is known for sure
  that $X$ is no more a \maybeeta. If $\sigma\hoUnifPb_i$ is in \llambda, then 
  % each variable that the
  % compiler used to replace a ``problematic'' term is assigned and hence
  by \cref{def:progBetaLL,def:progBetaRH} the associated \linkbeta
  is solved and removed. Note that, in this case, all
  calls to \Ue{} are between terms in \llambda and by \cref{th:fidelity-beta}
  fidelity is guaranteed. If $\sigma\hoUnifPb$ is still in \notllambda,
  then by \cref{def:progBetaFail} unification fails as the
  corresponding unification in \Fo{} would.
\end{proof}
